---
title: "STA199 Final Project"
subtitle: "Proposal"
format: html
editor: visual
---

```{r}
#| label: load-pkgs
#| message: false

library(tidyverse)
```

# Data 1

## Introduction and data

-   Identify the source of the data.

    -   Data was first collected in 1983 and has been collected every year since then. HUD used to use surveys of local housing markets to collect data, but now housing authorities conduct surveys on their behalf. They also accept telephone or mail surveys. 

-   State when and how it was originally collected (by the original data curator, not necessarily how you found the data).

    -   The Department of Housing and Urban Development\'s Office of Public Development and Research (HUD\'s PD&R)

-   Write a brief description of the observations.

    -   The data is divided by year then state and county. You can choose to view the whole state or just the county. The variables are year, state, counties, zip code, efficiency, one-bedroom, two-bedroom, three-bedroom, four-bedroom, and FMR percentile. 

## Research question

-   A well formulated research question. (You may include more than one research question if you want to receive feedback on different ideas for your project. However, one per data set is required.)

    -   How has the cost of rent changed post-COVID in North Carolina? 

-   A description of the research topic along with a concise statement of your hypotheses on this topic.

    -   The COVID-19 pandemic delivered a devastating blow to the economy, causing increased unemployment and inflation throughout the U.S. The cost of living has increased in multiple different states, and if we want to resolve this issue in North Carolina, we must understand the extent. By comparing North Carolina\'s data from 2019, 2020, 2021, and 2022, we will compare the changes in rent to test our hypothesis that rent increased Post-COVID and is making the cost of living unattainable. 

-   Identify the types of variables in your research question. Categorical? Quantitative?

    -   Year (2019, 2020, 2021, 2022), State (North Carolina), efficiency, one-bedroom, two-bedroom, three-bedroom, four-bedroom, and FMR percentile.

## Literature

-   Find one published credible article on the topic you are interested in researching.

    -    **<https://usafacts.org/articles/where-are-rents-rising-post-covid-19/>**

-   Provide a one paragraph summary about the article.

    -   This article does an overview of rent changes in the U.S and then discusses the county and zip codes with the highest rent increase. It looks closer at Arizona and San Francisco, finding that Maricopa and Pinal county In Arizona have the highest rent increase from 2020-2023, and that San Francisco have large rent differences by zip code.

-   In 1-2 sentences, explain how your research question builds on / is different than the article you have cited.

    -    Their visualizations give an example of what we are aiming for, but our research question will give a more in-depth look at the changes in North Carolina and which year had the highest increase.

## Glimpse of data

```{r}
#| label: fair-market-rent-data

FMR_data_2020 <- read_csv("data/FY20_FMR.csv")
FMR_data_2023 <- read_csv("data/FY23_FMR.csv")

glimpse(FMR_data_2020)
glimpse(FMR_data_2023)
```

# Data 2

## Introduction and data

-   Identify the source of the data.

    -   The United States Census Bureau has information regarding voter turnout and registration throughout the Current Population Survey from years 1994- 2020 focusing on the month of November. The data set is called CPS Voting Supplement in which one can select which years to view. 

-   State when and how it was originally collected (by the original data curator, not necessarily how you found the data).

-   Write a brief description of the observations.

## Research question

-   A well formulated research question. (You may include more than one research question if you want to receive feedback on different ideas for your project. However, one per data set is required.)

-   A description of the research topic along with a concise statement of your hypotheses on this topic.

-   Identify the types of variables in your research question. Categorical? Quantitative?

## Literature

-   Find one published credible article on the topic you are interested in researching.

-   Provide a one paragraph summary about the article.

-   In 1-2 sentences, explain how your research question builds on / is different than the article you have cited.

## Glimpse of data

```{r}
#| label: load-data-2

# add code here
```

# Data 3

## Introduction and data

-   Identify the source of the data.

    -   The data comes from NYC Slice, a dataset gathered by Liam Quigley, an independent New York City reporter.

-   State when and how it was originally collected (by the original data curator, not necessarily how you found the data).

    -   Liam Quigley himself over an 8-year period gathered data on 464 pizza slices throughout New York City.

-   Write a brief description of the observations.

    -   The variables included date, price, location, and style of pizza for 464 slices across New York City. Off first glance, it appears that price varied over the years (unsure whether or not there was a clear increase in price) and it seems that pepperoni pizzas generally tend to be more expensive.

## Research question

-   A well formulated research question. (You may include more than one research question if you want to receive feedback on different ideas for your project. However, one per data set is required.)

    -   Have pizza slices increased in price over time? How has the style of pizza impacted its price?

-   A description of the research topic along with a concise statement of your hypotheses on this topic.

    -   Data on prices of pizza slices can reveal certain trends about the prices of everyday goods for everyday citizens and how changes in the US's economy have impacted cost of living. With this being said, my hypothesis is that due to inflation, the price of pizza has clearly increased over the past 8 years.

-   Identify the types of variables in your research question. Categorical? Quantitative?

    -   Date, price, and style of pizza. The style of pizza is categorical, while the date and price are quantitative.

## Literature

-   Find one published credible article on the topic you are interested in researching.

    -   https://www.nytimes.com/2022/08/08/nyregion/inflation-nyc.html

-   Provide a one paragraph summary about the article.

    -   This New York Times article provides multiple anecdotes from case studies of everyday New Yorkers who struggle with rising food costs due to inflation. Stories of five different New Yorkers across multiple boroughs provide context to the difficulties provided by rising inflation. The article also describes how recent global events such as the pandemic and the war in Ukraine have impacted the prices of food products.

-   In 1-2 sentences, explain how your research question builds on / is different than the article you have cited.

    -   The New York Times article presents great anecdotal evidence to rising food prices, but this dataset would contribute a more objective perspective to a common New York City food staple: pizza.

## Glimpse of data

```{r}
#| label: load-nyc-slice-data

nyc_slice <- read_csv("data/nyc_slice_rawdata.csv")
glimpse(nyc_slice)
```
