---
title: "How Have Covid Change Rent Prices In North Carolina?"
subtitle: "Report"
format: 
  html:
    toc: false
editor: visual
execute:
  echo: true
---

# Introduction and Data:

For this research project, we will be investing: How has the cost of rent changed post-COVID in North Carolina? The COVID-19 pandemic delivered a huge blow to the economy, as a result, there has been a steady increase in inflation despite pay rates remaining roughly the same (Bernstein and Tedeschi). This has greatly increased the cost of living with rent prices skyrocketing in multiple states (Bernstein and Tedeschi). As college students, we want to analyze the extend of COVID-19's impact of rent prices in North Carolina, our home state. We will compare the rent prices of the one-bedroom, two-bedroom, three-bedroom, and four-bedroom for 2019 (pre-COVID) and 2022 (post-COVID). We hypothesized that rent prices from 2019 increased by 20% post-COVID.

We will use data collected by the Department of Housing and Urban Development's Office of Public Development and Research (HUD's PD&R). Data collection started in 1983 based on surveys of local housing markets. Now, housing authorities conduct surveys on their behalf, along with accepting telephone or mail surveys. Their full collection of data is divided by year, then state, and then county. There is also the option to view the full state or just the county. The variables included are zip code, efficiency, one-bedroom, two-bedroom, three-bedroom, four-bedroom, and Fair Market Rent (FMR) percentile. FMR is the amount an individual is likely to pay for rent in a particular area based on the term length listed by the landlord (HUD). The variable efficiency is defined as "the ratio of the rent estimate for Efficiency Units to the rent estimate \[certain bedroom units\]" (HUD). It is based on "the average of the past three estimates from the most recent American Community Survey 5-Year releases" (HUD). The variables, years, efficiency, one-bedroom, two-bedroom, three-bedroom, four-bedroom, and FMR percentile are quantitative variables. The variable state is a categorical variable.

Ethical concerns may involve the tenants in the property being disturbed by survey collectors or numerous telephone calls requesting for surveys. Another issue with privacy could be that tenants may not want everyone to know the rent prices of the area they live in. There is also an ethical concern regarding the areas and individuals being surveyed and how those results could alter rent prices alongside inflation.

# Literature Review:

This article does an overview of rent changes in the U.S. and then discusses the county and zip codes with the highest rent increase. It looks closer at Arizona and San Francisco, finding that Maricopa and Pinal counties in Arizona have the highest rent increase from 2020-2023 and that San Francisco has large rent differences by zip code. This literature also discusses in-depth the functions of FMR and how they measure final rent prices. This reading shows that COVID-19 increased rent prices but also had some outliers of counties with decreasing rent price changes.

# Methodology:

To test our hypothesis, we will use a linear regression model that looks how prices have changed in North Carolina as a result of COVID.

```{r, warning = FALSE, echo = FALSE, message = FALSE}
#| label: load data

library(tidyverse)
library(tidymodels)
library(knitr)

# MAKE SURE TO EXPLAIN WHAT DATES ARE INCLUDED IN THE 2020 and 2023 datasets.
rents_2020 <- read_csv("data/FY20_4050_FMRs_rev.csv")
rents_2021 <- read_csv("data/FY21_4050_FMRs_rev.csv")
rents_2022 <- read_csv("data/FY22_FMRs_revised.csv")
rents_2023 <- read_csv("data/FY23_FMRs_revised.csv")


# Modify_data function
# Parameters: Rent data frame, year of data, state to analyze
# Returns: A modified data frame with the following columns:
# state_alpha, year, countyname, bedrooms, fmr_0, fmr_1, fmr_2, fmr_3, fmr_4
modify_data <- function(data, inputYear, inputState) {
   data |>
    mutate(year = inputYear) |>
    filter(state_alpha == inputState) |>
    pivot_longer(
      cols = fmr_0:fmr_4, 
      names_to = "bedrooms",
      values_to = "fmr", 
      names_prefix = "fmr_"
    ) |>
    select(state_alpha, 
           year, 
           countyname,
           bedrooms, 
           fmr)
}

rents_2020 <- modify_data(rents_2020, 2020, "NC")
rents_2021 <- modify_data(rents_2021, 2021, "NC")
rents_2022 <- modify_data(rents_2022, 2022, "NC")
rents_2023 <- modify_data(rents_2023, 2023, "NC")


rents <- bind_rows(rents_2020, rents_2021, rents_2022, rents_2023)


```

### Did rent increase in your county after COVID-19?

**North Carolina counties where fair market rent increased from 2020 to 2023** [^1]

[^1]: **Source: HUD Fair Market Rent Data. Change in fair market rent is calculated by finding the average percent change of fair market rents from 2020 to 2023 of 0 to 4 bedroom accommodations.**

```{r, warning = FALSE, echo = FALSE, message = FALSE}
#| label: graph data
#| crop: true
#| fig-height: 4
#| fig-width: 12

# calculate percent change since 2020

rents <- rents |>
  group_by(countyname, bedrooms) |>
  arrange(year, .by_group = TRUE) |>
  mutate(fmr_pct = ((fmr - first(fmr)) / first(fmr)) * 100)

# calculate average change in fair market rent from 2020 to 2023 by county (irregardless of # bedrooms) 
nc_counties <- urbnmapr::counties |>
  filter(state_name == "North Carolina")
means <- rents |>
  group_by(year, countyname) |>
  summarize(mean_rent = mean(fmr_pct)) |>
  left_join(nc_counties, by = c("countyname"="county_name"))


# https://urban-institute.medium.com/how-to-create-state-and-county-maps-easily-in-r-577d29300bb2

means |>
  filter(year == 2023) |>
  filter(state_name == "North Carolina") |>
  ggplot(mapping = aes(long, lat, group = group, fill = mean_rent)) +
  geom_polygon(color = "#ffffff", size = .25) +
  coord_map(projection = "albers",  lat0 = 35, lat1 = -80) +
  theme(
        panel.grid.major = element_blank(),  # remove major grid lines
        panel.grid.minor = element_blank(),  # remove minor grid lines
        panel.border = element_blank(),  # remove panel border
        axis.line = element_blank(),  # remove axis lines
        axis.text = element_blank(),  # remove axis text labels
        axis.title = element_blank(), # remove axis titles
        axis.ticks = element_blank()  # remove tick marks on both axes
  ) +
  scale_fill_gradientn(guide = guide_colorbar(title.position = "top"),
                       colors = c("#1984c5", 
                                  "#63bff0", 
                                  "#a7d5ed", 
                                  "#e2e2e2",
                                  "#e1a692",
                                  "#de6e56", 
                                  "#e14b31", 
                                  "#c23728")) +
# Did rent increase in your county after COVID-19?
# Counties where one-bedroom fair market rent increased from 2020 to 2023, adjusted for inflation
  labs(fill = "Change in fair\nmarket rent (%)")

```

### Which counties in North Carolina had the largest rent increases after COVID-19?

**North Carolina counties with fair market rent increasing by 25% or more from 2020 to 2023**

```{r, warning = FALSE, echo = FALSE, message = FALSE}
#| label: report highest increase counties

high_increase <- means |>
  filter(year == 2023) |>
  distinct(countyname, mean_rent, .keep_all = TRUE) |>
  filter(mean_rent >= 25) |>
  ungroup() |>
  select(countyname, mean_rent) |>
  arrange(desc(mean_rent))

colnames(high_increase) <- c("County Name", "Change in Fair Market Price (%)")

high_increase |>
  kable()
```

### Which counties in North Carolina had the largest decrease in rents after COVID-19?

**North Carolina counties with fair market rent decreasing from 2020 to 2023**

```{r, warning = FALSE, echo = FALSE, message = FALSE}
#| label: report decrease counties

decreasing <- means |>
  filter(year == 2023) |>
  distinct(countyname, mean_rent, .keep_all = TRUE) |>
  filter(mean_rent <= 0) |>
  ungroup() |>
  select(countyname, mean_rent) |>
  arrange(mean_rent)

colnames(decreasing) <- c("County Name", "Change in Fair Market Price (%)")

decreasing |>
  kable()
```

```{r, warning = FALSE, echo = FALSE, message = FALSE}
#| label: box
#| fig-height: 8

rents |>
  ggplot(aes(y = year, x = fmr, group = year, fill = as.factor(year))) +
  geom_violin(show.legend = FALSE) +
  facet_wrap(facets = vars(bedrooms), nrow = 5) +
  labs(
    x = "40 Percentile Fair Market Rents",
    y = "Year",
    title = "Fair Market Rents from 2020 to 2023\nbased on the number of bedrooms"
  )
```

```{r}
rent_averages <- rents |>
  group_by(year, bedrooms) |>
  summarize(mean = mean(fmr)) |>
  mutate(year = year - 2020)

rent_averages |>
  kable()

regression <- linear_reg() |>
  set_engine("lm") |>
  fit(mean ~ year + bedrooms, data = rent_averages)

regression |>
  tidy() |>
  kable()

glance(regression)$r.squared
```

The linear regression model above can be used to find **estimate mean** fair market rents based on our data, if the trend in prices continues, even as the effects COVID-19 dwindle. From the linear regression model we can write an equation to estimate the mean fair market rents based on the number of years that have passed since 2020 and the number of bedrooms for the accommodation.

$$
\widehat{fmr} = (50.36 * years\_since\_2020) + 606.12 
$$

$$
+ (22.56 * one\_bedroom) + (177.24 * two\_bedroom) + (448.62 * three\_bedroom) + (680.97 * four\_bedroom)
$$

*where one_bedroom, two_bedroom, three_bedroom, and four_bedroom are Boolean values that represent if the fair market rent based on the number of bedrooms.*

Using the above linear regression model, we can find the estimate mean fair market rent (fmr) for a specified year and number of bedrooms. This prediction or estimate will be based on our data from 2020 through 2023 that we used in our data analysis.

For example, if we would like to predict the fair market rent of a 2 bedroom accommodation in the year 2024, we can use our linear regression model to predict the price assuming that the conditions from the years 2020 through 2023 hold:

$$
\widehat{fmr} = (50.36 * 4) + 606.12 + (177.24 * two\_bedroom)
$$ $$
\widehat{fmr} = (50.36 * 4) + 606.12 + (177.24 * 1)
$$ $$
\widehat{fmr} = 984.8
$$ In conclusion, from our linear regression model based on the mean FMRs from the years 2020 through 2023 in North Carolina, we can predict that a 2 bedroom accommodation in the year 2024 in North Carolina will have an estimate mean fair market rent of 984.8 dollars, holding all else constant.

# Results:

We started with understanding the rent increases per county after COVID-19. The results of the change in fair market rent % shows that numerous amounts of states increased their percentages. One state had a 40% change in FMR, and one state had a -10% change in FRM. Most states had a percent change of about 15-30% change. Our thesis estimated about a 20% increase post-COVID in which these results survey that the average change was about 25%.

We also researched which countries in NC has the largest rent increases after COVID-19. Based on our outcomes, the research indicates that Transylvania Country has the highest change in FMR, resting at 41.53% change. On the contrary we also calculated the countries in NC with the lowest rent increases after COVID-19. Camden County with -12.9% and Hyde County with -3.93 percent change. Furthermore, we decided to dive into the FMR from 2020-2023 based on the number of bedrooms. Based on the results, each year gradually, all housing, regardless of the number of rooms increased.

Our data analysis does not change the fair market rents increases or decreases to account for changes in inflation. We can improve on our analysis in the future by looking the fair market rents in North Carolina adjusted to account for inflation. Calculating the percent change in rents adjusted for inflation might allow us to make further conclusions, such as which counties in the state were the most impacted by COVID-19.

**Sources:**

https://www.huduser.gov/portal/datasets/fmr/fmrs/FY2023_code/2023bdrm_rent.odn?year=2023&cbsasub=METRO33860M33860&br_size=0

https://www.cbpp.org/research/poverty-and-inequality/tracking-the-covid-19-economys-effects-on-food-housing-and

https://www.huduser.gov/portal/datasets/fmr.html

https://www.whitehouse.gov/cea/written-materials/2021/04/12/pandemic-prices-assessing-inflation-in-the-months-and-years-ahead/
